---
title: "LINCS Signature MAE QC"
author: "Michael Tran"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

This notebook demonstrates the structure and basic use of the final
`MultiAssayExperiment` containing compound-level LINCS signatures, compound
annotations, and cell-line annotations.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(MultiAssayExperiment)
  library(SummarizedExperiment)
  library(S4Vectors)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(knitr)
  library(scales)
  library(uwot)
})

theme_set(theme_minimal(base_size = 12))

project_root <- normalizePath(
  if (dir.exists("data/results")) "." else "..",
  winslash = "/",
  mustWork = TRUE
)
knitr::opts_knit$set(root.dir = project_root)

mae_path <- file.path(
  project_root,
  "data",
  "results",
  "lincs_MultiAssayExperiment.rds"
)

if (!file.exists(mae_path)) {
  stop("MAE file not found: ", mae_path)
}

fmt_count <- function(x) {
  format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
}

fmt_number <- function(x, digits = 2) {
  format(round(x, digits), big.mark = ",", scientific = FALSE, trim = TRUE)
}

fmt_percent <- function(x, digits = 1) {
  paste0(format(round(100 * x, digits), nsmall = digits, trim = TRUE), "%")
}

is_present <- function(x) {
  !is.na(x) & nzchar(trimws(as.character(x)))
}

```

## Load the MAE

```{r load-data}
mae <- readRDS(mae_path)

assay_names <- names(experiments(mae))
signature_se <- experiments(mae)[["signatures"]]
signature_mat <- assay(signature_se)

compound_metadata <- as.data.frame(metadata(mae)$Drug.Metadata) |>
  rownames_to_column("Compound.Key") |>
  as_tibble()

coldata <- as.data.frame(colData(mae)) |>
  as_tibble()

feature_metadata <- as.data.frame(rowData(signature_se)) |>
  as_tibble()

selected_gene_metadata <- as.data.frame(metadata(mae)$Selected.Gene.Metadata) |>
  as_tibble()

cell_line_metadata <- if ("Cell.Line.Metadata" %in% names(metadata(mae))) {
  as.data.frame(metadata(mae)$Cell.Line.Metadata) |>
    rownames_to_column("Cell.Line.Key") |>
    as_tibble()
} else {
  tibble()
}

mae
```

## Dataset Overview

```{r dataset-overview, class.source="fold-hide"}
feature_type_counts <- feature_metadata |>
  count(LINCS.Feature.type, name = "features", sort = TRUE)

annotation_availability <- compound_metadata |>
  mutate(
    annotation_status = if_else(
      In.AnnotationDB,
      "AnnotationDB enriched",
      "No AnnotationDB match",
      missing = "No AnnotationDB match"
    )
  ) |>
  count(annotation_status, name = "compounds") |>
  mutate(
    fraction = compounds / sum(compounds),
    percent = scales::percent(fraction, accuracy = 0.1),
    label = paste0(
      annotation_status,
      "\n",
      scales::comma(compounds),
      " (",
      percent,
      ")"
    )
  )

assay_summary <- tibble(
  assay = assay_names,
  feature_type = "compound-level LINCS signature coefficients",
  assay_name = vapply(
    assay_names,
    \(assay_name) {
      paste(assayNames(experiments(mae)[[assay_name]]), collapse = ", ")
    },
    character(1)
  ),
  features = vapply(
    assay_names,
    \(assay_name) nrow(experiments(mae)[[assay_name]]),
    integer(1)
  ),
  compounds = vapply(
    assay_names,
    \(assay_name) ncol(experiments(mae)[[assay_name]]),
    integer(1)
  )
)

overview_tbl <- tibble(
  quantity = c(
    "Pipeline ID",
    "Pipeline version",
    "Experiments",
    "Assays",
    "Compounds in colData",
    "Drug metadata records",
    "Signature features",
    "Selected genes represented in metadata",
    "Median source signatures per compound",
    "Maximum source signatures per compound",
    "Compounds with PubChem CID",
    "Compounds without PubChem CID",
    "Cell line metadata records",
    "Cell lines in AnnotationDB"
  ),
  value = c(
    metadata(mae)$Pipeline$ID,
    metadata(mae)$Pipeline$Version,
    fmt_count(length(experiments(mae))),
    paste(assay_names, collapse = ", "),
    fmt_count(nrow(coldata)),
    fmt_count(nrow(compound_metadata)),
    fmt_count(nrow(signature_mat)),
    fmt_count(nrow(selected_gene_metadata)),
    fmt_count(stats::median(compound_metadata$LINCS.Signature.Count)),
    fmt_count(max(compound_metadata$LINCS.Signature.Count)),
    fmt_count(sum(!is.na(compound_metadata$Pubchem.CID))),
    fmt_count(sum(is.na(compound_metadata$Pubchem.CID))),
    fmt_count(nrow(cell_line_metadata)),
    fmt_count(
      if (nrow(cell_line_metadata) > 0) {
        sum(cell_line_metadata$In.AnnotationDB.CellLine, na.rm = TRUE)
      } else {
        0L
      }
    )
  )
)

kable(overview_tbl)
```

```{r assay-summary}
kable(assay_summary)
```

This MAE contains one assay, `signatures`. Rows are LINCS signature features and
columns are compounds keyed by `LINCS.CMap.Name`. The object is compound-level:
each assay column summarizes the source LINCS signatures associated with one
compound rather than one perturbation instance.

## Inspect the MAE Structure

The object layout is inspected directly below. The experiment container, assay
names, compound-level metadata, sample map, and object-level metadata can all be
accessed from the printed structure and the code shown in this section.

```{r object-contents}
mae

experiments(mae)

colnames(colData(mae))

head(as.data.frame(sampleMap(mae)))

names(metadata(mae))
```

The `sampleMap` is one-to-one for this dataset: both `primary` and `colname`
are `LINCS.CMap.Name`, and they point to the same compound columns in the
`signatures` assay.

## Compound Metadata

```{r compound-metadata-head}
compound_metadata |>
  select(
    LINCS.CMap.Name,
    Pubchem.CID,
    InChIKey,
    In.AnnotationDB,
    AnnotationDB.Name,
    LINCS.MOA,
    LINCS.Targets,
    LINCS.Signature.Count
  ) |>
  slice_head(n = 8) |>
  kable()
```

The main fields commonly used in downstream analysis are:

* `LINCS.CMap.Name`: the compound identifier used as the MAE primary key and as the assay column name.
* `Pubchem.CID`: the PubChem compound identifier used to join back to the base harmonized drug dataset.
* `InChIKey`: the structure-level identifier retained from source and AnnotationDB processing.
* `In.AnnotationDB`: whether the compound mapped to AnnotationDB metadata.
* `AnnotationDB.Name` and `AnnotationDB.SMILES`: the minimal AnnotationDB metadata carried with this dataset.
* `LINCS.MOA`, `LINCS.Targets`, and `LINCS.Aliases`: source LINCS compound annotations.
* `LINCS.Signature.Count`: the number of source LINCS signatures contributing to the compound-level signature.

```{r metadata-completeness}
metadata_completeness <- compound_metadata |>
  summarise(
    across(
      c(
        Pubchem.CID,
        InChIKey,
        AnnotationDB.Name,
        AnnotationDB.SMILES,
        LINCS.MOA,
        LINCS.Targets,
        LINCS.Aliases,
        AnnotationDB.Aliases
      ),
      \(value) sum(is_present(value))
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "field",
    values_to = "present"
  ) |>
  mutate(
    missing = nrow(compound_metadata) - present,
    percent_present = fmt_percent(present / nrow(compound_metadata))
  ) |>
  arrange(desc(present))

kable(metadata_completeness)
```

```{r annotationdb-coverage, fig.height=4.5, fig.width=7}
ggplot(
  annotation_availability,
  aes(x = "", y = fraction, fill = annotation_status)
) +
  geom_col(width = 1, color = "white") +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    lineheight = 0.95
  ) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#4C956C", "#C75C5C")) +
  labs(
    title = "AnnotationDB metadata availability for LINCS compounds",
    fill = NULL
  ) +
  theme_void()
```

## Cell Line Metadata

LINCS cell lines are represented as `cell_iname` feature rows in the
`signatures` assay rather than as MAE samples. Cell-line metadata is stored in
`metadata(mae)$Cell.Line.Metadata`, joined into the corresponding rowData rows,
and exported as `cell_line_metadata.tsv`.

```{r cell-line-metadata-summary}
cell_line_summary <- tibble(
  quantity = c(
    "Cell line metadata records",
    "Cell lines with Cellosaurus ID",
    "Cell lines in AnnotationDB",
    "Cell-line assay features"
  ),
  value = c(
    fmt_count(nrow(cell_line_metadata)),
    fmt_count(
      if (nrow(cell_line_metadata) > 0) {
        sum(is_present(cell_line_metadata$Cellosaurus.ID))
      } else {
        0L
      }
    ),
    fmt_count(
      if (nrow(cell_line_metadata) > 0) {
        sum(cell_line_metadata$In.AnnotationDB.CellLine, na.rm = TRUE)
      } else {
        0L
      }
    ),
    fmt_count(sum(feature_metadata$LINCS.Feature.type == "cell_iname"))
  )
)

kable(cell_line_summary)
```

```{r cell-line-metadata-preview}
cell_line_metadata |>
  select(
    any_of(
      c(
        "LINCS.Cell.Name",
        "Cellosaurus.ID",
        "In.AnnotationDB.CellLine",
        "Cell.Line.Name",
        "LINCS.CCLE.Name",
        "LINCS.Primary.Disease"
      )
    )
  ) |>
  slice_head(n = 8) |>
  kable()
```

## Signature Coverage

The source LINCS data can contain many signatures for the same compound. The
compound-level assay stores one signature column per compound while retaining
the number of source signatures in `LINCS.Signature.Count`.

```{r signature-count-summary}
signature_count_tbl <- compound_metadata |>
  summarise(
    compounds = n(),
    minimum = min(LINCS.Signature.Count),
    q1 = stats::quantile(LINCS.Signature.Count, 0.25),
    median = stats::median(LINCS.Signature.Count),
    mean = mean(LINCS.Signature.Count),
    q3 = stats::quantile(LINCS.Signature.Count, 0.75),
    maximum = max(LINCS.Signature.Count)
  ) |>
  mutate(across(everything(), fmt_number))

kable(signature_count_tbl)
```

```{r signature-count-histogram, fig.height=4}
axis_values <- c(1, 3, 10, 30, 100, 300, 1000, 3000)

ggplot(
  compound_metadata,
  aes(x = log10(LINCS.Signature.Count))
) +
  geom_histogram(bins = 45, fill = "#8EA8C3", color = "white") +
  scale_x_continuous(
    breaks = log10(axis_values),
    labels = scales::comma(axis_values)
  ) +
  labs(
    title = "Source signatures per compound",
    x = "Number of source signatures per compound",
    y = "Number of compounds"
  )
```

```{r most-replicated-compounds}
compound_metadata |>
  arrange(desc(LINCS.Signature.Count), LINCS.CMap.Name) |>
  select(
    LINCS.CMap.Name,
    Pubchem.CID,
    In.AnnotationDB,
    LINCS.MOA,
    LINCS.Targets,
    LINCS.Signature.Count
  ) |>
  slice_head(n = 12) |>
  kable()
```

## Compound Annotation Landscape

```{r moa-barplot, fig.height=6.5, fig.width=10}
moa_summary <- compound_metadata |>
  filter(is_present(LINCS.MOA)) |>
  count(LINCS.MOA, name = "compounds", sort = TRUE) |>
  slice_head(n = 15)

ggplot(
  moa_summary,
  aes(
    x = compounds,
    y = reorder(stringr::str_wrap(LINCS.MOA, width = 38), compounds)
  )
) +
  geom_col(fill = "#3B6EA8") +
  labs(
    title = "Most frequent LINCS mechanisms of action",
    x = "Number of compounds",
    y = "Mechanism of action"
  ) +
  theme(
    axis.text.y = element_text(size = 9),
    plot.margin = margin(5.5, 10, 5.5, 24)
  )
```

```{r target-barplot, fig.height=6, fig.width=10}
target_summary <- compound_metadata |>
  filter(is_present(LINCS.Targets)) |>
  count(LINCS.Targets, name = "compounds", sort = TRUE) |>
  slice_head(n = 15)

ggplot(
  target_summary,
  aes(
    x = compounds,
    y = reorder(stringr::str_wrap(LINCS.Targets, width = 38), compounds)
  )
) +
  geom_col(fill = "#6E9F64") +
  labs(
    title = "Most frequent LINCS targets",
    x = "Number of compounds",
    y = "Targets"
  ) +
  theme(
    axis.text.y = element_text(size = 9),
    plot.margin = margin(5.5, 10, 5.5, 24)
  )
```

## Signature Assay

```{r feature-metadata-head}
feature_metadata |>
  slice_head(n = 12) |>
  kable()
```

```{r feature-type-summary, fig.height=4}
kable(feature_type_counts)

ggplot(
  feature_type_counts,
  aes(x = reorder(LINCS.Feature.type, features), y = features)
) +
  geom_col(fill = "#7D5BA6") +
  labs(
    title = "Signature feature groups",
    x = "Feature group",
    y = "Number of features"
  ) +
  coord_flip()
```

The `signatures` assay stores numeric coefficients. Most entries are zero, so
the next summaries show both overall matrix completeness and non-zero
coefficient behavior.

```{r assay-quality-summary}
assay_values <- as.vector(signature_mat)
nonzero_values <- assay_values[is.finite(assay_values) & assay_values != 0]

assay_quality_tbl <- tibble(
  quantity = c(
    "Matrix rows",
    "Matrix columns",
    "Matrix entries",
    "Missing entries",
    "Non-finite entries",
    "Zero fraction",
    "Non-zero entries",
    "Minimum coefficient",
    "Median coefficient",
    "Mean coefficient",
    "Maximum coefficient"
  ),
  value = c(
    fmt_count(nrow(signature_mat)),
    fmt_count(ncol(signature_mat)),
    fmt_count(length(assay_values)),
    fmt_count(sum(is.na(assay_values))),
    fmt_count(sum(!is.finite(assay_values))),
    fmt_percent(mean(assay_values == 0, na.rm = TRUE)),
    fmt_count(length(nonzero_values)),
    fmt_number(min(assay_values, na.rm = TRUE), 4),
    fmt_number(stats::median(assay_values, na.rm = TRUE), 4),
    fmt_number(mean(assay_values, na.rm = TRUE), 4),
    fmt_number(max(assay_values, na.rm = TRUE), 4)
  )
)

kable(assay_quality_tbl)
```

```{r coefficient-histogram, fig.height=4}
tibble(log10_abs_coefficient = log10(abs(nonzero_values))) |>
  ggplot(aes(x = log10_abs_coefficient)) +
  geom_histogram(bins = 70, fill = "#C75C5C", color = "white") +
  labs(
    title = "Non-zero signature coefficient magnitudes",
    x = "log10 absolute coefficient",
    y = "Number of coefficients"
  )
```

```{r feature-variability}
feature_stats <- tibble(
  Feature.ID = rownames(signature_mat),
  Mean = rowMeans(signature_mat),
  SD = apply(signature_mat, 1, stats::sd),
  Zero.Fraction = rowMeans(signature_mat == 0)
) |>
  left_join(feature_metadata, by = "Feature.ID") |>
  arrange(desc(SD))

kable(
  feature_stats |>
    select(
      Feature.ID,
      LINCS.Feature.type,
      LINCS.Feature.value,
      Mean,
      SD,
      Zero.Fraction
    ) |>
    mutate(
      Mean = fmt_number(Mean, 4),
      SD = fmt_number(SD, 4),
      Zero.Fraction = fmt_percent(Zero.Fraction)
    ) |>
    slice_head(n = 15)
)
```

```{r feature-variability-plot, fig.height=6, fig.width=9}
top_feature_variability <- feature_stats |>
  slice_head(n = 20)

ggplot(
  top_feature_variability,
  aes(x = SD, y = reorder(Feature.ID, SD))
) +
  geom_col(fill = "#D49A3A") +
  labs(
    title = "Most variable signature features",
    x = "Standard deviation across compounds",
    y = "Feature"
  )
```

## UMAP View

A UMAP of the compound-by-feature matrix provides a compact view of global
signature structure. The plot below uses a reproducible sample of compounds and
colors points by the number of source signatures contributing to each
compound-level column.

```{r signature-umap, fig.height=5.5, fig.width=7}
set.seed(20260430)

variable_features <- feature_stats |>
  filter(is.finite(SD), SD > 0) |>
  pull(Feature.ID)

umap_sample_n <- min(5000L, ncol(signature_mat))

umap_compounds <- compound_metadata |>
  filter(LINCS.CMap.Name %in% colnames(signature_mat)) |>
  slice_sample(n = umap_sample_n)

umap_signatures <- signature_mat[
  variable_features,
  umap_compounds$LINCS.CMap.Name,
  drop = FALSE
]

umap_input <- scale(t(umap_signatures))
umap_input[is.na(umap_input)] <- 0

umap_layout <- uwot::umap(
  umap_input,
  n_neighbors = 15,
  min_dist = 0.2,
  metric = "cosine",
  n_threads = 1,
  n_sgd_threads = 1,
  verbose = FALSE
)

umap_tbl <- as_tibble(umap_layout, .name_repair = \(x) c("UMAP1", "UMAP2")) |>
  bind_cols(
    umap_compounds |>
      select(
        LINCS.CMap.Name,
        LINCS.Signature.Count,
        In.AnnotationDB
      )
  )

ggplot(
  umap_tbl,
  aes(
    x = UMAP1,
    y = UMAP2,
    color = LINCS.Signature.Count
  )
) +
  geom_point(alpha = 0.65, size = 0.8) +
  scale_color_gradientn(
    colors = c("#D8D8D8", "#3B6EA8", "#C75C5C"),
    trans = "log10",
    labels = scales::comma
  ) +
  labs(
    title = "UMAP of sampled LINCS compound signatures",
    subtitle = paste0(
      "Compound-level signatures for ",
      scales::comma(nrow(umap_tbl)),
      " sampled compounds."
    ),
    x = "UMAP 1",
    y = "UMAP 2",
    color = "Source\nsignatures"
  )
```

## High-Variance Feature Snapshot

This heatmap shows the most variable signature features across the compounds
with the largest number of contributing source signatures. Values are scaled by
feature for visualization.

```{r signature-heatmap, fig.height=7, fig.width=9}
top_heatmap_features <- feature_stats |>
  slice_head(n = 30) |>
  pull(Feature.ID)

top_heatmap_compounds <- compound_metadata |>
  arrange(desc(LINCS.Signature.Count), LINCS.CMap.Name) |>
  slice_head(n = 80) |>
  pull(LINCS.CMap.Name)

heatmap_mat <- signature_mat[
  top_heatmap_features,
  top_heatmap_compounds,
  drop = FALSE
]
heatmap_scaled <- t(scale(t(heatmap_mat)))
heatmap_scaled[!is.finite(heatmap_scaled)] <- 0

heatmap_tbl <- as.data.frame(heatmap_scaled) |>
  rownames_to_column("Feature.ID") |>
  as_tibble() |>
  pivot_longer(
    cols = -Feature.ID,
    names_to = "LINCS.CMap.Name",
    values_to = "Scaled.Value"
  ) |>
  mutate(
    Feature.ID = factor(Feature.ID, levels = rev(top_heatmap_features)),
    LINCS.CMap.Name = factor(
      LINCS.CMap.Name,
      levels = top_heatmap_compounds
    )
  )

ggplot(
  heatmap_tbl,
  aes(x = LINCS.CMap.Name, y = Feature.ID, fill = Scaled.Value)
) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#3B6EA8",
    mid = "white",
    high = "#C75C5C",
    midpoint = 0
  ) +
  labs(
    title = "Scaled high-variance signature features",
    x = "Compounds ordered by source signature count",
    y = "Feature",
    fill = "Scaled\nvalue"
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank()
  )
```

## Basic Access Patterns

The MAE can be used directly as an analysis-ready compound signature matrix.
Most analyses will either use the assay matrix in feature-by-compound form or
transpose it to compound-by-feature form for modeling.

```{r access-one-compound}
example_compound <- compound_metadata |>
  filter(In.AnnotationDB) |>
  arrange(desc(LINCS.Signature.Count), LINCS.CMap.Name) |>
  slice(1) |>
  pull(LINCS.CMap.Name)

example_metadata <- compound_metadata |>
  filter(LINCS.CMap.Name == example_compound) |>
  select(
    LINCS.CMap.Name,
    Pubchem.CID,
    InChIKey,
    AnnotationDB.Name,
    LINCS.MOA,
    LINCS.Targets,
    LINCS.Signature.Count
  )

kable(example_metadata)
```

```{r access-one-signature}
example_signature <- tibble(
  Feature.ID = rownames(signature_mat),
  coefficient = signature_mat[, example_compound]
) |>
  left_join(feature_metadata, by = "Feature.ID") |>
  arrange(desc(abs(coefficient))) |>
  select(
    Feature.ID,
    LINCS.Feature.type,
    LINCS.Feature.value,
    coefficient
  ) |>
  slice_head(n = 12)

kable(example_signature)
```

```{r access-target-subset}
egfr_compounds <- compound_metadata |>
  filter(grepl("EGFR", LINCS.Targets, fixed = TRUE)) |>
  pull(LINCS.CMap.Name)

egfr_signatures <- signature_mat[, egfr_compounds, drop = FALSE]

tibble(
  subset = "Compounds with EGFR in LINCS.Targets",
  compounds = length(egfr_compounds),
  features = nrow(egfr_signatures)
) |>
  kable()
```

```{r modeling-matrix}
modeling_matrix <- t(signature_mat)

tibble(
  object = c("feature_by_compound_matrix", "compound_by_feature_matrix"),
  rows = c(nrow(signature_mat), nrow(modeling_matrix)),
  columns = c(ncol(signature_mat), ncol(modeling_matrix))
) |>
  mutate(
    rows = fmt_count(rows),
    columns = fmt_count(columns)
  ) |>
  kable()
```

The compound-by-feature matrix is typically the convenient orientation for
machine learning or clustering workflows, while the feature-by-compound matrix
matches the stored `SummarizedExperiment` assay orientation.

## Session Info

```{r session-info}
sessionInfo()
```
